---
title: "Final Project"
author: "Chee Kay Cheong"
output: github_document
---

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(tidyverse)
library(caret)
library(glmnet)
library(randomForest)
```

# Introduction

Every year, the seasonal flu claims numerous lives across the globe. This highly contagious virus can spread easily and quickly, causing severe health complications, especially in
people with weakened immune systems. Vaccinations are essential to prevent illness, and if enough people receive them, they can create herd immunity, which can help protect the
wider community.

In the spring of 2009, an outbreak of H1N1 influenza, commonly known as "swine flu," spread worldwide, causing a death toll estimated to range from 151,000 to 575,000 in its first
year. A vaccine for the H1N1 flu virus became publicly available in October 2009. Later, in late 2009 and early 2010, the United States conducted the National 2009 H1N1 Flu Survey.
This phone survey collected personal information from participants and asked about their social, economic, and demographic background, views on disease and vaccine efficacy, and
steps taken to reduce the spread of infection. Gaining insights into the relationship between these characteristics and personal vaccination behavior can help inform future public
health initiatives.

### Research Question

The final project will employ a dataset acquired from [Kaggle](https://www.kaggle.com/datasets/arashnic/flu-data) that contains all the survey data gathered. We will use this
dataset **to identify the factors that may influence individuals' likelihood of receiving the H1N1 or seasonal flu vaccinations**.

### Rationale for research question

It is essential for public health to recognize the elements that affect people's probability to receive the H1N1 or seasonal flu shots for a variety of reasons:

1. **Improve disease prevention**
   + Understand what factors influence vaccine uptake can help public health officials target their efforts to increase vaccination rates and prevent outbreaks.
   
2. **Promote health equity**
   + Identifying factors that influence vaccine uptake can help address health disparities and ensure that everyone has access to vaccines. 
   + If certain groups are less likely to get vaccinated due to access barriers or mistrust of vaccines, public health officials can work to address these issues and increase vaccination rates in those populations.
   
3. **Reduce economic impact**
   + The flu can have a significant economic impact due to lost of productivity, healthcare costs, and other related expenses. Increasing vaccination rates and preventing outbreaks can help reduce these economic costs and benefit society as a whole.
   

# Data Preparation

```{r}

```





